Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-02-24 article-models-tt.tex
added empty context and empty families

\subsection{The basic judgments}
The judgments we wish to make are the standad six judgments for ``$\Gamma$ is a context'', ``$A$ is
a type in context $\Gamma$'' and ``$x$ is a term of $A$ in context $\Gamma$'' and the three equality
judgments coming with them.

\begin{align*}
\jctx*{\Gamma} & \jctxeq*{\Gamma}{\Gamma'}\\
\jtype*{\Gamma}{A} & \jtypeeq*{\Gamma}{A}{B}\\
\jterm*{\Gamma}{A}{x} & \jtermeq*{\Gamma}{A}{x}{y}.
\end{align*}

In contrast with standard practise, we don't assume that there is an empty context. The idea is that
if we interpret only the syntax of contexts, types and terms with extension, weakening and
substitution, we do not have a full model of type theory but we still have a 
structure resembling a category and categories need not have terminal objects. Therefore,
the rules that follow will describe only how to manipulate contexts, types and terms.
Of most of the inferences that we give there are two versions: one introducing an operation, the other asserting that the operation in question preserves judgmental equality.

\subsection{The basic rules for judgmental equality}
The rules for judgmental equality establish that it is an equivalence relation
in all three cases (contexts, types and terms).
\bgroup\small
\begin{align*}
& \inference{\jctx{\Gamma}}{\jctxeq{\Gamma}{\Gamma}} 
& & \inference{\jctxeq{\Gamma}{\Delta}}{\jctxeq{\Delta}{\Gamma}} 
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jctxeq{\Delta}{\greek{E}}}{\jctxeq{\Gamma}{\greek{E}}}\\
& \inference{\jtype{\Gamma}{A}}{\jtypeeq{\Gamma}{A}{A}} 
& & \inference{\jtypeeq{\Gamma}{A}{B}}{\jtypeeq{\Gamma}{B}{A}}
& & \inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtypeeq{\Gamma}{B}{C}}{\jtypeeq{\Gamma}{A}{C}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{x}{x}}
& & \inference{\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{A}{y}{x}}
& & \inference{\jtermeq{\Gamma}{A}{x}{y}\qquad\jtermeq{\Gamma}{A}{y}{z}}{\jtermeq{\Gamma}{A}{x}{z}}
\end{align*}
\egroup

@@

\begin{align*}
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtype{\Gamma}{A}}{\jtype{\Delta}{A}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtypeeq{\Gamma}{A}{B}}{\jtypeeq{\Delta}{A}{B}}\\
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jterm{\Gamma}{A}{x}}{\jterm{\Delta}{A}{x}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Delta}{A}{x}{y}}\\
& \inference{\jtypeeq{\Gamma}{A}{B}\qquad \jterm{\Gamma}{A}{x}}{\jterm{\Gamma}{B}{x}}
& & \inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{B}{x}{y}}
\end{align*}

@@

Note that by introducing extension on the level of types and families,
we introduce $\Sigma$-types at a very early stage.
However, we need substitution to make this precise.
\begin{align}
& \inference{\jtype{\Gamma}{A}}{\jctx{\ctxext{\Gamma}{A}}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtypeeq{\Gamma}{A}{B}}{\jctxeq{\ctxext{\Gamma}{A}}{\ctxext{\Delta}{B}}}\\
& \inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\ctxext{A}{P}}}
& & \inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtypeeq{\ctxext{\Gamma}{A}}{P}{Q}}{\jtypeeq{\Gamma}{\ctxext{A}{P}}{\ctxext{B}{Q}}}
\end{align}

\subsubsection{Extension is compatible with itself}
The inference rules asserting that extension is compatible with itself assert
that contexts are unstructured lists of type declarations. This rule is
unavoidable if we want that for a type $A$ in context $\Gamma$, a type
in context $A$ is the same thing as a type in context $\ctxext{\Gamma}{A}$.

\begin{align}
& \inference{\jtype{\Gamma}{A}\qquad\jtype{\ctxext{\Gamma}{A}}{P}}
  {\jctxeq{\ctxext{{\Gamma}{A}}{P}}{\ctxext{\Gamma}{{A}{P}}}}\\
& \inference{\jtype{\ctxext{\Gamma}{A}}{P}\qquad\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}
  {\jtypeeq{\Gamma}{\ctxext{{A}{P}}{Q}}{\ctxext{A}{{P}{Q}}}}
\end{align}

@@

In the following inference rules we assume $\jctx{\Gamma}$ and in the rules
asserting a judgmental equality we furthermore assume $\jctxeq{\Gamma}{\Gamma'}$.
\begin{align}
& \inference{\jctx{\Delta}}{\jtype{\Gamma}{\ctxwk{\Gamma}{\Delta}}} 
& & \inference{\jctxeq{\Delta}{\Delta'}}{\jtypeeq{\Gamma}{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma'}{\Delta'}}}\\
& \inference{\jtype{\Delta}{B}}{\jtype{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}}
& & \inference{\jtypeeq{\Delta}{B}{B'}}{\jtypeeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma'}{B'}}}\\
& \inference{\jterm{\Delta}{B}{y}}{\jterm{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma}{y}}} 
& & \inference{\jtermeq{\Delta}{B}{y}{y'}}{\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma}{y}}{\ctxwk{\Gamma'}{y'}}} 
\end{align}

@@

 In the following inference rules we assume that $\jtype{\Gamma}{A}$ and in the
 rules asserting a judgmental equality we assume furthermore that 
 $\jtypeeq{\Gamma}{A}{A'}$.
\begin{align}
& \inference{\jtype{\Gamma}{B}}{\jtype{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}}
& & \inference{\jtypeeq{\Gamma}{B}{B'}}{\jtypeeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A'}{B'}}}\\
& \inference{\jtype{\ctxext{\Gamma}{B}}{Q}}
{\jtype{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}}
& & \inference{\jtypeeq{\ctxext{\Gamma}{B}}{Q}{Q'}}
{\jtypeeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A'}{Q'}}}\\
& \inference{\jterm{\Gamma}{B}{y}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}}
& & \inference{\jtermeq{\Gamma}{B}{y}{y'}}{\jtermeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}{\ctxwk{A'}{y'}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jterm{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}}
& & \inference{\jtermeq{\ctxext{\Gamma}{B}}{Q}{g}{g'}}
{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}{\ctxwk{A'}{g'}}}
\end{align}

\subsubsection{Weakening is compatible with extension}

@@

\begin{align}
& \inference
  {\jtype{\Delta}{B}}
  {\jtypeeq{\Gamma}{\ctxwk{\Gamma}{\ctxext{\Delta}{B}}}{\ctxext{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma}{B}}}}\\
& \inference
  {\jtype{\Delta}{B}\qquad\jtype{\ctxext{\Delta}{B}}{Q}}
  {\jtypeeq
    {\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}
    {\ctxwk{\Gamma}{\ctxext{B}{Q}}}
    {\ctxext{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma}{Q}}}}
\end{align}

In the following inference rules we assume that $\jtype{\Gamma}{A}$.

\begin{align}
& \inference
  {\jtype{\ctxext{\Gamma}{B}}{Q}}
  {\jtypeeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\ctxext{B}{Q}}}{\ctxext{\ctxwk{A}{B}}{\ctxwk{A}{Q}}}}\\
& \inference
  {\jtype{\ctxext{{\Gamma}{B}}{Q}}{R}}
  {\jtypeeq
    {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk{A}{\ctxext{Q}{R}}}
    {\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}}}
\end{align}

\subsubsection{Weakening is compatible with itself}
To conclude the rules for weakening, we state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two contexts $\Gamma$ and $\Delta$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jctx{\blank} \ar{r}{\greek{E}\mapsto\ctxwk{\Delta}{E}} \ar{d}[swap]{\greek{E}\mapsto\ctxwk{\Gamma}{E}} & \jtype{\Delta}{\blank} \ar{d}{B\mapsto\ctxwk{\Gamma}{B}}\\
\jtype{\Gamma}{\blank} \ar{r}[swap]{A\mapsto\ctxwk{{\Gamma}{\Delta}}{A}} & \jtype{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\blank}
\end{tikzcd}
\end{equation*}
There is also a version of this diagram in which all happens in a context. Thus,
we get two sets of inference rules. For weakening by contexts we get:
\begin{align}
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}\qquad\jctx{\greek{E}}}
          {\jtypeeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{{\Delta}{\greek{E}}}}
            {\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{\greek{E}}}}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}\qquad\jtype{\greek{E}}{C}}
          {\jtypeeq{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{\greek{E}}}}}
            {\ctxwk{\Gamma}{{\Delta}{C}}}
            {\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{C}}}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}\qquad\jterm{\greek{E}}{C}{t}}
          {\jtermeq{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{\greek{E}}}}}
            {\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{C}}}{\ctxwk{\Gamma}{{\Delta}{t}}}{\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{t}}}}
\end{align}
For weakening by types we get:
\begin{align}
& \inference{\jtype{\Gamma}{A}\qquad\jtype{\Gamma}{B}\qquad\jtype{\Gamma}{C}}
          {\jtypeeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{{B}{C}}}
            {\ctxwk{{A}{B}}{{A}{C}}}}\\
& \inference{\jtype{\Gamma}{A}\qquad\jtype{\Gamma}{B}\qquad\jtype{\ctxext{\Gamma}{C}}{R}}
          {\jtypeeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\\
& \inference{\jtype{\Gamma}{A}\qquad\jtype{\Gamma}{B}\qquad\jterm{\Gamma}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
\end{align}

>>>>

Much of the rules we state are just compatibility rules of extension, weakening
and substitution with each other. To get the overview, we list the sections
where these compatibility rules are described in the following table (in this
table, the
subsection mentioned in row $X$ and column $Y$ consideres the rules of the
operation $Y\circ X$):

\begin{center}
\begin{tabular}{r|ccc}
& extension & weakening & substitution\\
\hline
extension & \autoref{comp-ee} & \autoref{comp-ew} & \\
weakening & \autoref{comp-we} & \autoref{comp-ww} & \autoref{comp-ws}\\
substitution & \autoref{comp-se} & \autoref{comp-sw} & \autoref{comp-ss}
\end{tabular}
\end{center}

Since extension by a context acts trivially on families and terms there are
no rules added for extension followed by substitution.

\subsection{The basic judgments}
The type theory we describe here is a theory of contexts, families of
contexts and terms thereof. The families of contexts are by some authors called
dependent contexts, but they are handled a bit differently here because they
become the primary object of study. Dependent contexts can be types; they could
be seen as atomic or indecomposable dependent contexts.

Thus we make eight types of judgments in our type theory: ``$\Gamma$ is a context'',
``$A$ is a family of contexts over $\Gamma$'', ``$A$ is a type in context $\Gamma$''
and ``$x$ is a term of the family $A$ of contexts over $\Gamma$''. The other four
judgments are for judgmental equality.

\begin{align*}
\jctx*{\Gamma} & \jctxeq*{\Gamma}{\Gamma'}\\
\jfam*{\Gamma}{A} & \jfameq*{\Gamma}{A}{B}\\
\jtype*{\Gamma}{A} & \jtypeeq*{\Gamma}{A}{B}\\
\jterm*{\Gamma}{A}{x} & \jtermeq*{\Gamma}{A}{x}{y}.
\end{align*}

If $A$ is a type
in context $\Gamma$, then $A$ is also a family of contexts over $\Gamma$. Being
a term of type $A$ means the same thing as being a term of the family $A$ of contexts.
Two types in context $\Gamma$ are judgmentally equal precisely when they are equal
as context families. Moreover, if a family $B$ of contexts over $\Gamma$ is
judgmentally equal to a type $A$ in context $\Gamma$, then $B$ is a type in
context $\Gamma$. This is expressed by the following four inference rules:

\begin{align*}
& \inference{\jfam{\Gamma}{A}}{\jfam{\Gamma}{A}} & & \inference{\jfameq{\Gamma}{A}{B}}{\jfameq{\Gamma}{A}{B}}\\
& \inference{\jfam{\Gamma}{A}\quad\jfameq{\Gamma}{A}{B}}{\jfam{\Gamma}{B}}
& & \inference{\jfam{\Gamma}{A}\quad\jfameq{\Gamma}{A}{B}}{\jfameq{\Gamma}{A}{B}}
\end{align*}

\subsection{The basic rules for judgmental equality}
The rules for judgmental equality establish that it is an equivalence relation
in all three cases (contexts, types and terms).
\bgroup\small
\begin{align*}
& \inference{\jctx{\Gamma}}{\jctxeq{\Gamma}{\Gamma}} 
& & \inference{\jctxeq{\Gamma}{\Delta}}{\jctxeq{\Delta}{\Gamma}} 
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jctxeq{\Delta}{\greek{E}}}{\jctxeq{\Gamma}{\greek{E}}}\\
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{A}{A}} 
& & \inference{\jfameq{\Gamma}{A}{B}}{\jfameq{\Gamma}{B}{A}}
& & \inference{\jfameq{\Gamma}{A}{B}\qquad\jfameq{\Gamma}{B}{C}}{\jfameq{\Gamma}{A}{C}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{x}{x}}
& & \inference{\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{A}{y}{x}}
& & \inference{\jtermeq{\Gamma}{A}{x}{y}\qquad\jtermeq{\Gamma}{A}{y}{z}}{\jtermeq{\Gamma}{A}{x}{z}}
\end{align*}
\egroup

@@

\begin{align*}
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jfam{\Gamma}{A}}{\jfam{\Delta}{A}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jfameq{\Gamma}{A}{B}}{\jfameq{\Delta}{A}{B}}\\
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jterm{\Gamma}{A}{x}}{\jterm{\Delta}{A}{x}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Delta}{A}{x}{y}}\\
& \inference{\jfameq{\Gamma}{A}{B}\qquad \jterm{\Gamma}{A}{x}}{\jterm{\Gamma}{B}{x}}
& & \inference{\jfameq{\Gamma}{A}{B}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{B}{x}{y}}
\end{align*}

\subsection{The empty context}
There is an empty context and over any context there is an empty family of
contexts. We do not assume that the empty context is a type, that would be like
assuming that the multiplicative unit of a ring is prime. The empty family
always has a unique term.

\begin{align}
& \inference{}{\jctx{\emptyc}}\\
& \inference{\jctx{\Gamma}}{\jfam{\Gamma}{\emptyf\Gamma}}\\
& \inference{\jctx{\Gamma}}{\jterm{\Gamma}{\emptyf[\Gamma]}{\emptytm[\Gamma]}}\\
& \inference{\jterm{\Gamma}{\emptyf[\Gamma]}{x}}{\jtermeq{\Gamma}{\emptyf[\Gamma]}{x}{\emptytm[\Gamma]}}
\end{align}

Moreover, if $\Gamma$ is a context family over the
empty context, then $\Gamma$ is a context and every context is a context
family over the empty context. Note that this allows us to speak
of terms of contexts too.

\begin{align}
& \inference{\jctx{\Gamma}}{\jfam{\emptyc}{\Gamma}} 
& & \inference{\jfam{\emptyc}{\Gamma}}{\jctx{\Gamma}}\\
& \inference{\jctxeq{\Gamma}{\Delta}}{\jfameq{\emptyc}{\Gamma}{\Delta}}
& & \inference{\jfameq{\emptyc}{\Gamma}{\Delta}}{\jctxeq{\Gamma}{\Delta}}
\end{align}

\subsubsection{The empty context is compatible with itslef}
The empty context $\emptyc$ may be considered as a family of contexts over the empty
context. When we do this, we get $\emptyf[\emptyc]$.
\begin{equation}
\inference{}{\jfameq{\emptyc}{\emptyc}{\emptyf[\emptyc]}}
\end{equation}
In the future, we shall denote $\emptyf[\Gamma]$ by $\emptyf$. The above rule
guarantees that this will not cause confusion. Likewise, we shall denote
$\emptytm[\Gamma]$ by $\emptytm$.

@@

Note that by introducing extension on the level of types and families,
we introduce $\Sigma$-types at a very early stage.
However, we need substitution to make this precise.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jctx{\ctxext{\Gamma}{A}}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jfameq{\Gamma}{A}{B}}{\jctxeq{\ctxext{\Gamma}{A}}{\ctxext{\Delta}{B}}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfam{\Gamma}{\ctxext{A}{P}}}
& & \inference{\jfameq{\Gamma}{A}{B}\qquad\jfameq{\ctxext{\Gamma}{A}}{P}{Q}}{\jfameq{\Gamma}{\ctxext{A}{P}}{\ctxext{B}{Q}}}
\end{align}

\subsubsection{Extension is compatible with the empty context}
The following rule asserts that extension by $\emptyc$ leaves the contexts unchanged.
\begin{align}
& \inference{\jctx{\Gamma}}{\jctxeq{\ctxext{\emptyc}{\Gamma}}{\Gamma}}\\
& \inference{\jctx{\Gamma}}{\jctxeq{\ctxext{\Gamma}{\emptyf}}{\Gamma}}\\
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\ctxext{\emptyf}{A}}{A}}
\end{align}

\subsubsection{Extension is compatible with itself}\label{comp-ee}
The inference rules asserting that extension is compatible with itself assert
that contexts are unstructured lists of type declarations. This rule is
unavoidable if we want that for a type $A$ in context $\Gamma$, a type
in context $A$ is the same thing as a type in context $\ctxext{\Gamma}{A}$.

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\ctxext{\Gamma}{A}}{P}}
  {\jctxeq{\ctxext{{\Gamma}{A}}{P}}{\ctxext{\Gamma}{{A}{P}}}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}\qquad\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}}
  {\jfameq{\Gamma}{\ctxext{{A}{P}}{Q}}{\ctxext{A}{{P}{Q}}}}
\end{align}

@@

In the following inference rules we assume $\jctx{\Gamma}$ and in the rules
asserting a judgmental equality we furthermore assume $\jctxeq{\Gamma}{\Gamma'}$.
\begin{align}
& \inference{\jctx{\Delta}}{\jfam{\Gamma}{\ctxwk{\Gamma}{\Delta}}} 
& & \inference{\jctxeq{\Delta}{\Delta'}}{\jfameq{\Gamma}{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma'}{\Delta'}}}\\
& \inference{\jfam{\Delta}{B}}{\jfam{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}}
& & \inference{\jfameq{\Delta}{B}{B'}}{\jfameq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma'}{B'}}}\\
& \inference{\jterm{\Delta}{B}{y}}{\jterm{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma}{y}}} 
& & \inference{\jtermeq{\Delta}{B}{y}{y'}}{\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma}{y}}{\ctxwk{\Gamma'}{y'}}} 
\end{align}

@@

 In the following inference rules we assume that $\jfam{\Gamma}{A}$ and in the
 rules asserting a judgmental equality we assume furthermore that 
 $\jfameq{\Gamma}{A}{A'}$.
\begin{align}
& \inference{\jfam{\Gamma}{B}}{\jfam{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}}
& & \inference{\jfameq{\Gamma}{B}{B'}}{\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A'}{B'}}}\\
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
{\jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}}
& & \inference{\jfameq{\ctxext{\Gamma}{B}}{Q}{Q'}}
{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A'}{Q'}}}\\
& \inference{\jterm{\Gamma}{B}{y}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}}
& & \inference{\jtermeq{\Gamma}{B}{y}{y'}}{\jtermeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}{\ctxwk{A'}{y'}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jterm{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}}
& & \inference{\jtermeq{\ctxext{\Gamma}{B}}{Q}{g}{g'}}
{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}{\ctxwk{A'}{g'}}}
\end{align}

\subsubsection{Weakening is compatible with the empty context}

Weakening by the empty context $\emptyc$ leaves contexts, context families and terms
unchanged:
\begin{align}
& \inference{\jctx{\Gamma}}{\jctxeq{\ctxwk{\emptyc}{\Gamma}}{\Gamma}}\\
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\ctxwk{\emptyc}{A}}{A}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\ctxwk{\emptyc}{x}}{x}}
\end{align}

Likewise, weakening by the empty family $\emptyf$ over a context $\Gamma$ leaves families, families over that and
terms of those unchanged:
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\ctxwk{\emptyf}{A}}{A}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{\emptyf}{P}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\ctxwk{\emptyf}{f}}{f}}
\end{align}

\subsubsection{Weakening is compatible with extension}\label{comp-we}

@@

\begin{align}
& \inference
  {\jfam{\Delta}{B}}
  {\jfameq{\Gamma}{\ctxwk{\Gamma}{\ctxext{\Delta}{B}}}{\ctxext{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma}{B}}}}\\
& \inference
  {\jfam{\Delta}{B}\qquad\jfam{\ctxext{\Delta}{B}}{Q}}
  {\jfameq
    {\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}
    {\ctxwk{\Gamma}{\ctxext{B}{Q}}}
    {\ctxext{\ctxwk{\Gamma}{B}}{\ctxwk{\Gamma}{Q}}}}
\end{align}

In the following inference rules we assume that $\jfam{\Gamma}{A}$.

\begin{align}
& \inference
  {\jfam{\ctxext{\Gamma}{B}}{Q}}
  {\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\ctxext{B}{Q}}}{\ctxext{\ctxwk{A}{B}}{\ctxwk{A}{Q}}}}\\
& \inference
  {\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}
  {\jfameq
    {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk{A}{\ctxext{Q}{R}}}
    {\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}}}
\end{align}

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two contexts $\Gamma$ and $\Delta$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jctx{\blank} \ar{r}{\greek{E}\mapsto\ctxwk{\Delta}{E}} \ar{d}[swap]{\greek{E}\mapsto\ctxwk{\Gamma}{E}} & \jfam{\Delta}{\blank} \ar{d}{B\mapsto\ctxwk{\Gamma}{B}}\\
\jfam{\Gamma}{\blank} \ar{r}[swap]{A\mapsto\ctxwk{{\Gamma}{\Delta}}{A}} & \jfam{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\blank}
\end{tikzcd}
\end{equation*}
There is also a version of this diagram in which all happens in a context. Thus,
we get two sets of inference rules. For weakening by contexts we get:
\begin{align}
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}\qquad\jctx{\greek{E}}}
          {\jfameq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{{\Delta}{\greek{E}}}}
            {\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{\greek{E}}}}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}\qquad\jfam{\greek{E}}{C}}
          {\jfameq{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{\greek{E}}}}}
            {\ctxwk{\Gamma}{{\Delta}{C}}}
            {\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{C}}}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}\qquad\jterm{\greek{E}}{C}{t}}
          {\jtermeq{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{\greek{E}}}}}
            {\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{C}}}{\ctxwk{\Gamma}{{\Delta}{t}}}{\ctxwk{{\Gamma}{\Delta}}{{\Gamma}{t}}}}
\end{align}
For weakening by types we get:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\Gamma}{C}}
          {\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{{B}{C}}}
            {\ctxwk{{A}{B}}{{A}{C}}}}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\ctxext{\Gamma}{C}}{R}}
          {\jfameq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\Gamma}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
\end{align}

\subsubsection{The empty context is compatible with weakening}
The following rules express that when the empty context or context family is
weakened, the result is the empty context family.

\begin{align}
& \inference{\jctx{\Gamma}}{\jfameq{\Gamma}{\ctxwk{\Gamma}{\emptyc}}{\emptyf}}\\
& \inference{\jfam{\Gamma}{A}}{\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\emptyf}}{\emptyf}}
\end{align}

\subsubsection{Extension is compatible with weakening}\label{comp-ew}
The rules expressing that extension is compatible with weakening assert that
weakening by an extension is the same thing as weakening twice in the
appropriate way.

In the following, we assume that $\jctx{\Gamma}$ and $\jfam{\Gamma}{A}$.

\begin{align}
& \inference{\jctx{\Delta}}
    {\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{\ctxext{\Gamma}{A}}{\Delta}}{\ctxwk{A}{{\Gamma}{\Delta}}}}\\
& \inference{\jfam{\Delta}{B}}
    {\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{{\Gamma}{\Delta}}}}{\ctxwk{\ctxext{\Gamma}{A}}{B}}{\ctxwk{A}{{\Gamma}{B}}}}\\
& \inference{\jterm{\Delta}{B}{g}}
    {\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{{\Gamma}{\Delta}}}}{\ctxwk{A}{{\Gamma}{B}}}{\ctxwk{\ctxext{\Gamma}{A}}{g}}{\ctxwk{A}{{\Gamma}{g}}}}
\end{align}

The rules for weakening by a type and a family over it are similar. We assume that
$\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$.
